---
title: "HGDP clustering"
output: html_document
date: "2024-12-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## HGDP data set 

Here we use data from the Human Genome Diversity Project (HGDP)
https://www.internationalgenome.org/data-portal/data-collection/hgdpit to run 
through a full population genetic pipeline from raw data to population structure
analysis and visualisation, to create the plots seen in the Paper <>.

We downloaded the data from <???>

## Data preparation

Load necessary packages:
```{r}
# load packages 
library(tidygenclust)
library(tidypopgen)
library(readr)
```

Load the data into `gen_tibble` object:
```{r}
# path to PLINK
bed_path <- "../../hgdp_zhe_1_12-10-2024/hgdp650.qc.hg19.bed"

# load the data into gentibble using tidypopgen
hgdp_gt <- gen_tibble(bed_path, backingfile = "../../hgdp_zhe_1_12-10-2024/gentibble/hgdp_raw") 

# load the meta data
hgdp_meta <- read.table("../../hgdp_zhe_1_12-10-2024/HGDPid_populations.txt", header = TRUE)
```

Attach meta data to `gen_tibble`:
```{r}
# attach meta data to gentibble based on id 
hgdp_gt <- hgdp_gt %>% mutate(
  population = hgdp_meta$population[match(hgdp_gt$id, hgdp_meta$Id)],region = hgdp_meta$Region[match(hgdp_gt$id, hgdp_meta$Id)]
)
```


## Data preparation 

Remove more than 5% missing as suggested in ADMIXTURE manual:
```{r}
hgdp_gt <- hgdp_gt %>% select_loci_if(loci_missingness(genotypes)<0.05)
```

Remove monomorphic loci:
```{r}
hgdp_gt <- hgdp_gt %>% select_loci_if(loci_maf(genotypes)>0)
```

Update backing file:
```{r}
hgdp_gt <- gt_update_backingfile(hgdp_gt, backingfile = "../../hgdp_zhe_1_12-10-2024/gentibble/hgdp_filt")
gt_save(hgdp_gt, "../../hgdp_zhe_1_12-10-2024/gentibble/hgdp_filt")
```


Impute any remaining missing values:
```{r}
# impute the data
hgdp_gt <- hgdp_gt %>% gt_impute_simple(method = "random")

# check
hgdp_gt %>% gt_has_imputed()
```

LD prune the data:
```{r}
# LD prune the data based on the threshold stated in ADMIXTURE manual "plink --bfile rawData --indep-pairwise 50 10 0.1"
hgdp_ldprune_gt <- hgdp_gt %>% select_loci_if(loci_ld_clump(genotypes, thr_r2 = 0.1, size = 50, use_positions = FALSE))

#check remaining
hgdp_gt %>% count_loci()
hgdp_ldprune_gt %>% count_loci()
```

Update backing file and impute again:
```{r}
#update backing file 
hgdp_ldprune_gt <- gt_update_backingfile(hgdp_ldprune_gt, backingfile = "../../hgdp_zhe_1_12-10-2024/gentibble/hgdp_ld")

#save
gt_save(hgdp_ldprune_gt, "../../hgdp_zhe_1_12-10-2024/gentibble/hgdp_ld")
```

```{r}
hgdp_ldprune_gt <-gt_load("../../hgdp_zhe_1_12-10-2024/gentibble/hgdp_ld")
# impute the data
hgdp_ldprune_gt <- hgdp_ldprune_gt %>% gt_impute_simple(method = "random")
```




Check data with PCA:
```{r}
# PCA
hgdp_pca <- hgdp_ldprune_gt %>% gt_pca_partialSVD()

library(ggplot2)
autoplot(hgdp_pca, type = "scores") +
  aes(color = hgdp_ldprune_gt$region) +
  labs(color = "Region")
```

## Population structure analysis - ADMIXTURE `tidypopgen`

Run ADMIXTURE through `tidypopgen`:
```{r}
# group by population for later plotting
hgdp_ldprune_gt <- hgdp_ldprune_gt %>% group_by(population)

# run ADMIXTURE for k = 2-15 and 1 repeat to get an idea of the best k 
hgdp_adm_cv <- hgdp_ldprune_gt %>% 
  gt_admixture(k = 2:15, n_runs = 1, crossval = TRUE, 
               n_cores = 2, seed = 123, conda_env = "none")
```

Check `gt_admix` results object:
```{r}
# summarise the admixture results
hgdp_adm_cv %>% summary()
```

Plot the cross-validation:
```{r}
# plot the CV
hgdp_adm_cv %>% autoplot()
```


Organise meta data
```{r}
# split bantu and rename some populations
# if Bantu then combine population and Geographic_origin
hgdp_meta$population[hgdp_meta$population=="Bantu"] <-paste( hgdp_meta$population[hgdp_meta$population=="Bantu"],
hgdp_meta$Geographic_origin[hgdp_meta$population=="Bantu"], sep="_")
# replace inappropriate labels
hgdp_meta$population <- gsub("Pygmies","HG", hgdp_meta$population)
# create a factor with 7 regions and order them geographically
hgdp_meta$Pop7Groups <- sub("Est", "East", hgdp_meta$Pop7Groups)
hgdp_meta$Pop7Groups <- factor(hgdp_meta$Pop7Groups, levels = c("Africa", "Middle_East", "Europe", "Central_South_Asia", "East_Asia", "Oceania","America"))
hgdp_meta <- hgdp_meta %>% arrange(Pop7Groups, population)
pop_levels <- unique(hgdp_meta$population)
```

Reorder admix
```{r}
hgdp_admix <- readRDS("../../hgdp_zhe_1_12-10-2024/clustering/hgdp_admixture_pop.rds")
hgdp_meta  <- hgdp_meta %>% filter(.data$Id %in% hgdp_admix$id)
hgdp_meta <- hgdp_meta[match(hgdp_admix$id, hgdp_meta$Id),]
pops <- hgdp_meta$population
pops <- factor(pops, levels = pop_levels)

#reorder gt_admix
hgdp_admix <- gt_admix_reorder_q(hgdp_admix, pops)
```

Barplot for single run:
```{r}
# plot barplot for given k and run using the groupings inherited from the grouped tibble
hgdp_admix %>% autoplot(type = "barplot", k = 5, run = 1)    
```

Save `gt_admix`:
```{r}
# save results 
hgdp_adm_cv %>% saveRDS("../../hgdp_zhe_1_12-10-2024/clustering/hgdp_admixture.rds")
```


##  Population structure analysis - fastmixture `tidygenclust`

```{r}
# run fastmixture for k = 2:5 and 5 runs -> or however many want to showcase clumppling visualisations
hgdp_fastm <- hgdp_ldprune_gt %>% 
  gt_fastmixture(k = c(2:5), n_repeats = 5, threads = 2,
                 seed = c(123,234,345,456,567))

#or load in previous
hgdp_fastm <- readRDS("../../hgdp_zhe_1_12-10-2024/clustering/hgdp_fastm.rds")
```

Reorder admix
```{r}
hgdp_meta  <- hgdp_meta %>% filter(.data$Id %in% hgdp_fastm$id)
hgdp_meta <- hgdp_meta[match(hgdp_fastm$id, hgdp_meta$Id),]
pops <- hgdp_meta$population
pops <- factor(pops, levels = pop_levels)

#reorder gt_admix
hgdp_fastm <- gt_admix_reorder_q(hgdp_fastm, pops)
```


Check `gt_admix` results object:
```{r}
# summarise the fastmixture results
hgdp_fastm %>% summary()
```

Barplot for single run:
```{r}
# plot barplot for given k and run using the groupings inherited from the grouped tibble:
hgdp_fastm %>% autoplot(type = "barplot", k = 5, run = 1)
```

Save `gt_admix`:
```{r}
# save results
hgdp_fastm %>% saveRDS("../../hgdp_data/clustering/hgdp_fastmixture.rds")
```


## clumppling visualisations in `tidygenclust`

Run clumpppling
```{r}
# run clumppling fastmixture
hgdp_fastm_clump <- hgdp_fastm %>% gt_clumppling()
```

Create ordered vector for meta
```{r}
clump_pops <- as.vector(pops)
#arrange pops2 by pop_levels
clump_pops <- clump_pops[order(match(clump_pops, pop_levels))]
```

All modes plot
```{r}
hgdp_fastm_clump %>% autoplot(type = "modes", group = clump_pops)
```

Major modes plot
```{r}
# plot all modes with grouping annotation
hgdp_fastm_clump %>% autoplot(type = "major_modes", group = clump_pops)
```



